---
title: "p8105_hw2_ps3395.Rmd"
author: "PENG"
date: "2023-10-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE)
```

```{r libraries}
library(tidyverse)
library(readxl)
```

# Problem 2

### Read and clean data from *"202207 Trash Wheel Collection Data.xlsx"*

```{r}
#Read and clean the Mr. Trash Wheel sheet
Mr_trash = 
  readxl::read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 1 ,range = "A2:N550") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(homes_powered = round ((weight_tons * 500) / 30, 0),
         trashID = "Mr")

#Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda
professor_trash = 
  readxl::read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 2 ,range = "A2:M97") |>
  janitor::clean_names() |>
  drop_na(dumpster)|>
  mutate(homes_powered = round ((weight_tons * 500) / 30, 0),
         trashID = "professor")

gwynnda_trash = 
  readxl::read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 4 ,range = "A2:K110") |>
  janitor::clean_names() |>
  drop_na(dumpster)|>
  mutate(homes_powered = round ((weight_tons * 500) / 30, 0),
         trashID = "gwynnda")

#finding same columns in three dataset
common_col = Reduce(intersect, list(names(Mr_trash), names(professor_trash), names(gwynnda_trash)))

#merge three dataset by same columns to a single tidy data
merge_trash = merge(Mr_trash, professor_trash, by= common_col,all = TRUE) 
merge_trash = merge(merge_trash, gwynnda_trash, by = common_col, all = TRUE)|>
  janitor::clean_names()
```

#### Description 

Three datasets are involved in Problem 2, including *"Mr."*, *"Professor"* and *"Gwynnda"* Trash Wheel. Which containing `r nrow(Mr_trash)`, `r nrow(professor_trash)`, and `r nrow(gwynnda_trash)` observations respectively. Moreover, there are `r length(common_col)` common variables involved in these three datasets, such as

* `month`, `year` and `date` for recording the time.
* `weight_tons` and `volume_cubic_yards` represent the weights and volumes of the trash.
* `plastic_bottles`, `polystyrene` and `cigarette_butts` indicate trash amount in different types.
* `homes_powered` represents the average time that electricity converted from garbage can be used by a household.

Then merging these three datasets into one dataframe called *"merge_trash"*, which contains `r nrow(merge_trash)` observations and `r ncol(merge_trash)` variables. \newline

For available data, the total weight of trash collected by Professor Trash Wheel was `r sum(pull(professor_trash, weight_tons))` tons, and the total number of cigarette butts collected by Gwynnda in July of 2021 was `r format(sum(pull(filter(gwynnda_trash, month == "July" , year == 2021),cigarette_butts)), big.mark = ",")`.

